|

### 2024-08-22 reproducibility/manuscript/manuscript.qmd
docs(manuscript): update methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

  Single-cell RNA Velocity has dramatically advanced our ability to model

@@

RNA Velocity is a powerful computational framework to estimate the time
derivative of gene expression
[@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj]. The framework
has been used to study developmental cell lineage trajectories over a broad
spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj],
pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors
differentiation [@Bergen2020-pj]. Yet, recent state-of-the-art RNA Velocity
approaches such as velocyto and scVelo [@La_Manno2018-lj] have several
limitations. First, they may predict developmental trajectories that do not
exist in nature, as it has been shown in the analysis of fully mature blood
cells [@Zheng2017-bz;@Bergen2021-qz]. Second, data preprocessing steps (e.g.,
selection of top variable genes, dimensionality reduction) affect velocity
estimates that are difficult to characterize and might lead to artifactual
predictions. In addition, non-linear spliced and unspliced imputation procedures
could distort phase portrait geometry and recovered trajectories
([@fig-model]a). Third, most RNA velocity algorithms perform estimates
independently for each gene, and therefore cannot exploit the natural
interdependence of gene expression. Fourth, a unified description of the
temporal relation between cells can only be assigned based on postprocessing
steps that aggregate heuristically the gene-wise fits into a shared latent time
([@fig-model]a). Finally, current RNA velocity frameworks only output point
estimates of the velocity vector field and do not provide uncertainty
estimation. Importantly, the absence of statistical confidence estimates
increases false discoveries as it leaves users without a way to evaluate the
robustness of the methods' outputs. To overcome these limitations, we developed
a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates
RNA dynamics and infers cell trajectories using Pyro [@Bingham2018-id], a robust
and efficient probabilistic programming language.

@@

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

>>>>

  Single-cell RNA velocity has dramatically advanced our ability to model

@@

RNA velocity is a powerful computational framework to estimate the time
derivative of gene expression
[@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj]. The framework
has been used to study developmental cell lineage trajectories over a broad
spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj],
pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors
differentiation [@Bergen2020-pj]. Yet, recent state-of-the-art RNA Velocity
approaches such as velocyto and scVelo [@La_Manno2018-lj] have several
limitations. First, they may predict developmental trajectories that do not
exist in nature, as it has been shown in the analysis of fully mature blood
cells [@Zheng2017-bz;@Bergen2021-qz]. Second, data preprocessing steps (e.g.,
selection of top variable genes, dimensionality reduction) affect velocity
estimates that are difficult to characterize and might lead to artifactual
predictions. In addition, non-linear spliced and unspliced imputation procedures
could distort phase portrait geometry and recovered trajectories
([@fig-model]a). Third, most RNA velocity algorithms perform estimates
independently for each gene, and therefore cannot exploit the natural
interdependence of gene expression. Fourth, a unified description of the
temporal relation between cells can only be assigned based on postprocessing
steps that aggregate heuristically the gene-wise fits into a shared latent time
([@fig-model]a). Finally, current RNA velocity frameworks only output point
estimates of the velocity vector field and do not provide uncertainty
estimation. Importantly, the absence of statistical confidence estimates
increases false discoveries as it leaves users without a way to evaluate the
robustness of the methods' outputs. To overcome these limitations, we developed
a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates
RNA dynamics and infers cell trajectories using Pyro [@Bingham2018-id], a robust
and efficient probabilistic programming language.

@@

We assume the dynamical gene expression is determined by the RNA splicing
process, and infer the unspliced and spliced gene expression level from the
differential equations proposed in velocyto [@La_Manno2018-lj] and scVelo
[@Bergen2020-pj]
\begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right),
   \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)
  -\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}
In the equation, the subscript $c$ is the cell dimension, $g$ is the gene dimension,
$\left( u\left( \tau^{(k_{cg})} \right), s\left( \tau^{(k_{cg})} \right) \right)$ 
are the unspliced and spliced expression functions given the
change of time per cell and gene. $\tau_{cg}$ represents the displacement of time per
cell and gene with 
\begin{align}
 \tau^{(k_{cg})} &= \operatorname{softplus} \left( t_{c} - {t_{0}^{(k_{cg})}}_g \right) \\
 & = \log( 1 + \exp (t_c - {t_{0}^{(k_{cg})}}_g)), 
\end{align}
in which $t_c$ is the shared time per cell, ${t_{0}^{(kcg)}}_g$ is the
gene-specific switching time. Each cell and gene combination has its
transcriptional state $k_{cg} \in \{ 0, 1 \}$, where $0$ indicates the
activation state and $1$ indicates the expression state. Each gene has two
switching times for representing activation and repression: ${t_{0}^{(0)}}_g$ is
the first switching time corresponding to when the gene expression starts to be
activated, ${t_0^{(1)}}_g$  is the second switching time corresponding to when
the gene expression starts to be repressed. We note that $\alpha^{(1)}$ is
shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
each gene.

The analytic solution of the differential equations to predict
spliced and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
[@Bergen2020-pj;@Li2021-qa] and given in
Eqs. \ref{eq-solution-u}-\ref{eq-solution-s2}.
\begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}

To simplify these equations, consider the case when $k_{cg} = 0$ and 
$\beta_g \neq \gamma_g$. Then, 
\begin{align}
u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
  \label{eq-sol-usimp} \\
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
  \nonumber\\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
\end{align}
When $k_{cg} = 0$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(0)}\right)$ 
has the same solution, and $s\left(\tau^{(0)}\right)$ becomes
\begin{align}
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
\end{align}
When $k_{cg} = 1$ and $\beta_g \neq \gamma_g$, then
\begin{align}
u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
\end{align}
When $k_{cg} = 1$ and $\beta_g = \gamma_g$, then $u\left(\tau^{(1)}\right)$ 
has the same solution, and $s\left(\tau^{(1)}\right)$ becomes
\begin{align}
s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
\end{align}

## Variational inference {#sec-methods-inference}

Given observations $\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)$,
we would like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align}
but exact Bayesian inference is intractable in this model. We use Pyro to
automatically integrate out the local discrete latent variables $k$, which is
defined as the cell and gene transcriptional state (see above), and approximate
the posterior over the remaining latent variables using variational inference
[@Bingham2018-id;@Kucukelbir2016-bk], which converts intractable integrals into
optimization problems that can be solved with off-the-shelf tools. In
variational inference, we maximize a tractable bound (the evidence lower bound
(ELBO)) on the Kullback-Leibler (KL) divergence
$$\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)$$
between a parametric family of tractable probability distributions q and the
intractable true posterior
\begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align}
We approximate our model’s posterior distribution $p( \theta, \psi \vert
\tilde{X})$ with a tractable family of probability distributions
$q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)$, where
$q_{\phi_1}(\theta)$ is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
$q_{\phi_2}(\psi)$ is a multivariate Gaussian distribution with a learnable
location parameter and a learnable low-rank covariance matrix. We solve the
resulting stochastic optimization problem using a version of Automatic
Differentiation Variational Inference (ADVI) [@Kucukelbir2016-bk]: we obtain
gradient estimates by differentiating a Monte Carlo estimate of the ELBO and
update the variational parameters  using stochastic gradient ascent.  Our
implementation and experiments use generic variational families and Monte Carlo
ELBO and gradient estimators provided by Pyro [@Bingham2018-id], an open source
library for probabilistic machine learning, and a version built into Pyro of the
adaptive stochastic gradient ascent algorithm Adam augmented with gradient
clipping to enhance numerical stability during training.

## Model training {#sec-methods-training}

For the pancreas, PBMC, and uni-fate and bi-fate LARRY single-cell data, since
the data dimension is relatively small, we run Model 1 and Model 2 with a
minimum of 100 epochs and a maximum of 4000 epochs. For each epoch, we input
spliced and unspliced read counts from all the cells and determine the
convergence condition with an early stopping strategy in which the patience is
set to 45 and consumes one patience if the minimal ELBO improvement of training
data per epoch is lower than $10^{-4}$ of previous loss. The learning rate is
set to be $10^{-2}$ with a decay rate of $0.11/4000$  per epoch. For the
multi-fate LARRY dataset, since this is a large dataset with over $4 \times
10^4$ cells, we use mini batches of cells to train Model 1 and Model 2 with a
minimum of 100 epochs and a maximum of 1000 epochs. Specifically, the batch size
was set to 4000 cells for both models with an early stopping patience 45 and
consume one patience if the minimal ELBO improvement of training data per epoch
lower than $10^{-3}$ of previous loss. The learning rate is set to be $10^{-2}$
with a decay rate of $0.11/1000$  per epoch. All models were trained on a
machine with an NVIDIA A100 GPU and the CentOS 7 operating system.
